---
title: "Week 1 Lab"
subtitle: "Part 2: All about R"
author: "Connor Wiegand"
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    embed-resources: true
    # html-math-method: mathjax
  #  revealjs:
  #   theme: serif
  #   preview-links: true
  #   chalkboard: true
  #   slide-number: true
  #   menu:
  #       side: right
  #   scrollable: true
  #   smaller: true
  #   incremental: true
  #   slide-level: 3

editor:
   render-on-save: true
    
---

# All about R

## A little about R

### Assignment
We can assign a name to different objects in R, using the notation `LHS <- RHS`^[(Left/Right) Hnad Side]. This way, if RHS is long and complicated, or is the result of another computation or program, we can just say `LHS`, and the computer will know to insert `RHS`. For instance, we can assign x to be the number 2 by running `x <- 2`. To view (aka `print`) x, we simply call it:

```{r ex1}
x <- 2
x
```

### Data Types
There are many types of data in R. They can be accessed by passing the data to the `class()` function. Let's review a few key types:

```{r ex2}
class(x)
class(1+1)
```

```{r ex3}
class("string")
foo <- "bar"
class(foo)
```

---

A vector in R is of the form `c(a,b,c,d)`. All of the components of the vector must have the same data type. 

Moreover, "vector" is not a data type in R. The components of the vector determine it's type. 
```{r ex4}
my_vec <- c(1,2,4,7)
class(my_vec)
```

---

```{r}
mat <- matrix(1:4, nrow = 2)
mat
class(mat)
```

```{r}
# A hash mark let's us write comments in our code
my_df <- as.data.frame(mat)
class(my_df)
```

## Packages
R’s functionality can be extended through packages, which are collections of functions, data sets, and documentation bundled together. 

In order to use a package in R, two things have to be true:

1. The package is installed on your computer

2. You have to load the package into the R _session_
    - I.e., if you close R, you likely have to re-load the package. 


### Ol' Reliable
The `base` way of managing packages in R is to install them using 
`install.packages('<package name>')` and load them using `library(<package name>)`. 

:::{.callout-note collapse="true"}
# `<Notation>`
By the way, you won't really need to put anything in <> at any point. So, if my name is Connor Wiegand, and I read some form of `<your name>` in a tutorial in this class, I should replace all of `<your name>` with `Connor Wiegand`. That is, 

- `<your name>` = `Connor Wiegand`
- `'<your name>'` = `'Connor Wiegand'`
- `"<your name>"` = `"Connor Wiegand"`
:::

So, if I wanted to use the `dplyr` package for the first time, I would start by running `install.packages("dplyr")`. Then, when I want to use `dplyr` in my session, I run `library(dplyr)`.

::: callout-tip
You only need to install a package **once**! Then it is on your system. Next time, you don't have to run `install.packages()`, you can just call `library()`. The step of calling `library(<pkg name>)` must be re-done every time you close RStudio. 
:::

### Pacman

`pacman` is a package manager that has been around for a long time. The `pacman` package in R comes with a number of tools for package management, with the star-pupil being `p_load`. To use this command, you must first install the pacman package the usual way:
```{r eval=F}
install.packages("pacman")
```

Assume we have loaded pacman via `library(pacman)`, but have not loaded or installed any other packages. Then, if I wanted to install _and_ load `dplyr` in one step, I would just run `pacman(dplyr)`. In fact, if I wanted to load 3 commonly used packages at once, I could run 
```{r eval=F}
pacman(dplyr, magrittr, ggplot2)
```
And all of my packagees would be installed (if they are not already) and loaded into my current session.

::: {.callout-warning collapse=T}
There are times when installing via `pacman` does some funky things under the hood of your machine. When installing key packages for the first time, I typically use `install.packages`.
:::

---

### `<pkg>::<fn>`

In RStudio, typing `<package name>` followed by two colons will pull up a set of available functions from the `<package name>` package. Moreover, if `<pkg-A>` is installed on your machine, you can use `pkg-A::fn()` to access the `fn` function, _regardless of whether or not `pkg-A` is loaded_. That means that, after installing `pacman` for the first time, you can use the syntax
```{r eval=F}
pacman::p_load(pkg-A, pkg-B, etc.)
```
throughout your various programs to load in a package. 