---
title: "Lab 5: Heteroskedasticity"
# subtitle: "Part 2"
author: Connor Wiegand (adapted from Andrew Dickinson)
format:
  html:
    theme: 
      - flatly
      - ../lab-style.scss
    toc: true
    toc-depth: 3
    execute: 
      # eval: false
      warning: false
      error: false
    embed-resources: true
    # html-math-method: mathjax
knitr: 
   opts_chunk: 
      class-output: code-out
      class-error: err-out
      class-message: msg-out
# editor:
#    render-on-save: true
---

# DGP

```{r}
library(tidyverse)

set.seed(12345)
n=1000
data = tibble(
  x=rnorm(n, 2, 1),
  group=c(rep(0,n/2), rep(1,n/2)),
  error = rnorm(n, 0, 1),
  y_homo =  error,
  y_hetero1= error*x,
  y_hetero2= error*abs(x-2),
  y_hetero3= error*(group+1)*3,  
) 
# Optional: Save the data to a CSV file
#write_csv(data, file="Data_Heteroskedasticity.csv")
```

### Visualizing the Generated Data


Visualization is a crucial step in understanding the data and the relationships among variables. Let's visualize the relationships between $x$ and different simulated dependent variables using `ggplot2`.

::: {.panel-tabset}
## `y_homo`

```{r}
ggplot(data=data, aes(x=x, y=y_homo)) +
  geom_point()
```

## `y_hetero1`

```{r}
ggplot(data=data, aes(x=x, y=y_hetero1)) +
  geom_point()
```

## `y_hetero2`

```{r}
ggplot(data=data, aes(x=x, y=y_hetero2)) +
  geom_point()
```

## `y_hetero3`

```{r}
ggplot(data=data, aes(x=x, y=y_hetero3, color=factor(group))) +
  geom_point(alpha=0.9) +
  scale_color_manual(
      values = viridisLite::viridis(2)
   ) +
  labs(
    color = "Group"
  )
  theme(
    legend.position = "bottom"
  )
```

:::

# Goldfeld-Quandt Test

The Goldfeld-Quandt (GQ) test is a classical statistical procedure used to check for heteroskedasticity in a regression model. Heteroskedasticity, a violation of one of the Gauss-Markov assumptions, occurs when the variance of the error terms is not constant across observations. This test is particularly useful when the variances are believed to change at a particular point of a covariate or across groups. The [GQ test]{.hi} splits the data into different segments, performs separate regression analyses on these segments, and tests whether the variances of the residuals from these regressions are significantly different.

## Goldfeld-Quandt Test in R

1. **Sort the Data**:
    - Arrange the data in ascending order based on the independent variable suspected to exhibit heteroskedasticity.
    
```r
data = data %>%
  arrange(x)
```

2. **Split the Data**:
    - Divide the data into two or more segments. A common approach is to split it into thirds and discard the middle third.
    
```r
first_third = nrow(data)/3
third_third = floor((nrow(data)/3) * 2) + 2
data_1 = data[1:first_third, ]
data_2 = data[third_third:n, ]
```

3. **Run Separate Regressions**:
    - Perform regression analyses on these segments separately.
    
```r
fit1 = lm(y_hetero2 ~ x, data_1 )  %>% summary()
fit2 = lm(y_hetero2 ~ x, data_2 )  %>% summary()
```

4. **Compute Residual Sum of Squares (RSS)**:
    - Calculate the sum of squared residuals for each regression.
    
```r
rss1 = sum(resid(fit1)^2)
rss2 = sum(resid(fit2)^2)
```

5. **Conduct Hypothesis Testing**:
    - Compare the variances of the residuals from these regressions using an F-test.
    - The null hypothesis ($H_0$) is that the variances are equal (Homoskedasticity), against the alternative hypothesis ($H_1$) that they are not equal (Heteroskedasticity).
    
```r
tstat = rss2/rss1
pf(q = tstat, df1 = 333-1, df2 = 333-1, lower.tail = F)
```

The *GQ test* is a straightforward and intuitive method to test for heteroskedasticity, especially when there is a prior belief that the variance of errors may change at a particular point of a covariate.

# White's Test

White's Test is a statistical procedure that helps detect heteroskedasticity in a regression model. Unlike the Goldfeld-Quandt Test, White's Test does not require the data to be sorted, nor does it require the assumption that variances change at a particular point of a covariate. Instead, it's a general test for heteroskedasticity without specifying the form of heteroskedasticity.

## White's Test in R

1. **Run the Initial Regression**:
    - Perform a regression analysis using the suspected model.
    
```r
fit = lm(y_hetero2 ~ x, data)
```

2. **Compute Squared Residuals**:
    - Obtain the residuals from the regression and square them.
    
```r
data = data %>% mutate(resid = resid(fit)^2)
```

3. **Regress Squared Residuals on Original Independent Variable(s), Their Squares and Cross Products**:
    - Perform a regression of the squared residuals on the original independent variable(s), their squares, and the cross products of the independent variables.
    
```r
fit = lm(resid ~ x + I(x^2), data) %>% summary()
```

4. **Conduct Hypothesis Testing**:
    - The test statistic is computed as $N \cdot R^2$ from the regression in step 3, where $N$ is the sample size and $R^2$ is the coefficient of determination from the regression.
    - Under the null hypothesis of homoskedasticity, the test statistic follows a chi-square distribution with degrees of freedom equal to the number of regressors (excluding the constant term) in the regression in step 3.

```r
teststat = fit$r.squared * n
pchisq(q = teststat, df = 2, lower.tail = F)
```

White's Test provides a non-specific approach to detect heteroskedasticity, making it a flexible option when the form of heteroskedasticity is unknown. It's a widely used test due to its generality and simplicity in implementation.